---
title: "Individual project pre analysis"
author: "Claire Campion"
date: "5/11/2022"
output: pdf_document
bibliography: bib3.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Overview

Bumble bee populations are increasingly under threat from habitat fragmentation,
pesticides, pathogens, and climate change. Climate is likely a prime driver of 
declines in abundance and distribution, as bumble bees are limited by their 
thermal tolerance [@kerr2015]. However, the tolerance of whole organisms often
exceeds that of their gametes; for example, insects can be sterilized by 
temperatures below their upper thermal tolerances [@van_heerwaarden2021; 
@david2005]. In bumble bees, males are independent from the colony and can 
withstand extreme temperatures, but whether these temperatures compromise 
spermatozoa is unclear. Using commercially-reared bumble bee (*Bombus
impatiens*) males, we measured how exposure to sublethal temperatures near male 
critical thermal minimum and maximum impact spermatozoa viability. We measured 
temperature effects on spermatozoa in intact males to determine if males are 
potentially protecting spermatozoa. A LIVE/DEAD™ Cell Imaging Kit (Invitrogen™)
and cell counter (Nexcelom Cellometer Spectrum Image Cytometry System) were used
to estimate sperm count and viability in males exposed to 45°C for 85 minutes 
and 4°C for 85 min and 48 hours. Cells were stained with LIVE/DEAD kit
immediately after temperature exposures and were compared to control males 
which were held at room temperature (22°C) throughout experiment.The purpose of 
this study is to determine how exposure time may influence temperature effects 
on viability. This gap is important to address as bumble bee populations
continue to decline and potential sterility to males could be a contributing
factor. 


My goal for this project is to be able to create box plots and violin graphs 
to visualize these results and begin to understand the effects of temperature 
on the reproductive fitness in bumble bee males. I hope to begin to get these 
figures in a place where that can be published!

# 2. Dataset
The dataset I will be using for this experiment include cell counts of male 
spermatozoa. The data consists of live cells, dead cells, total cells, and 
percent viability(live/total). Other information included in the datasets are 
treatment, date, colony, species, date, mass, beeid, image number, and sex. I 
get my data from a automated microscope and I take 3 images of each sample and
average them to get a accurate calculation of the cell count. I have 2 datasets,
one for cold and one for heat treatment as explained above. 

I anticipate there not being many issues with this data but I want to explore
some addition aspects I could add to my boxplot such as changing the y axis to
log scale or creating a histogram to see if differences are due to mass or other
variables. 



```{r packages}
library(tidyverse, message== FALSE)
library(readxl, message== FALSE)
library(ggplot2, message== FALSE)
library(patchwork, message== FALSE)


```

```{r datasets}
chillvia <- read_xlsx("chillcoma exp.xlsx")
str(chillvia)
heatvia <- read_xlsx("heat exp.xlsx") 
str(heatvia)
```

# 3. Tidying Data
In this section, my goal is to tidy data, make sure all columns are correct 
data type, removing any potential NAs, calculating the mean live, dead, and 
total counts for each unique beeID, and grouping by bee ID. 

```{r tidy datasets}
#For the chill dataset, I using the mutate function to assign each column is 
#the correct data type.  I want my treatment as a factor and the rest as 
#numeric value.
chillvia <- chillvia %>%
  mutate(treat=factor(treat), 
         time=as.numeric(time),
         live=as.numeric(live), 
         dead=as.numeric(dead), 
         total=as.numeric(total), 
         pc_viability=as.numeric(pc_viability))

#Next I am grouping the dataset by BeeID and using summarize to calculate the 
#means for viability, live, dead, total cell counts, mass, and time. I also am
#taking the unique value of each beeIDs treatment and colony. 
cvia1 <- chillvia %>% 
  group_by(beeid) %>%
  summarize(via = mean(pc_viability, na.rm=TRUE), 
            live = mean(live, na.rm=TRUE), 
            dead = mean(dead, na.rm=TRUE),
            total= live+dead,
            treat = unique(treat),
            time = mean(time, na.rm=T), 
            mass = mean(mass, na.rm=TRUE),
            colony = unique(colony))
cvia1

#For the heat dataset, I using the mutate function to assign each column is the 
#correct data type. I want my treatment as a factor and the rest as numeric
#values.
heatvia <- heatvia %>%
  mutate(treat=factor(treat), 
         time=as.numeric(time),
         live=as.numeric(live), 
         dead=as.numeric(dead), 
         total=as.numeric(total), 
         pc_viability=as.numeric(pc_viability))

#Next I am grouping the dataset by BeeID and using summarize to calculate the 
#means for viability, live, dead, total cell counts, mass, and time. I also am 
#taking the unique value of each beeIDs treatment and colony. 
hvia1 <-  heatvia %>% 
  group_by(beeid) %>%
  summarize(via = mean(pc_viability, na.rm=TRUE), 
            live = mean(live, na.rm=TRUE), 
            dead = mean(dead, na.rm=TRUE),
            total= live+dead,
            treat = unique(treat),
            time = mean(time, na.rm=T), 
            mass = mean(mass, na.rm=TRUE),
            colony = unique(colony))
hvia1

#renaming variables to be more decriptive for graphs 

cvia1<-cvia1%>% 
  mutate(treat=ifelse(treat=="Control", "Control", ifelse(treat=="Chillcoma", "Cold", treat)))
         
hvia1<-hvia1%>% 
  mutate(treat=ifelse(treat=="control", "Control", ifelse(treat=="heat", "Heat",ifelse(treat=="hs", "Heat Stupor", treat))))



```
# 4. Scatterplots of mass
Using scatterplot, I will visualize how mass varies among treatments in both
datasets. This is to ensure any results are not due to mass differences. 

```{r scatterplots of mass}
#comparing mass in cold treatment experiments by total cell count and % 
#viability

cvia1 %>%
  ggplot(aes(x=mass, y=total, col=treat)) +
  geom_point()

cvia1 %>%
  ggplot(aes(x=mass, y=via, col=treat)) +
  geom_point()


#comparing mass in heat treatment experiments by total cell count and % 
#viability

hvia1 %>%
  ggplot(aes(x=mass, y=total, col=treat)) +
  geom_point()

hvia1 %>%
  ggplot(aes(x=mass, y=via, col=treat)) +
  geom_point()


```
This graph is showing that mass and viability/ cell count are likely not
correlated. This is important to consider before moving forward so we know that
a decrease in viability is not due to mass of organism. 


# 5. Plotting Viability in both datasets with boxplots
Using boxplots, I will visualize how temperature impacts spermatozoa viability
in male bumble bees. The first box plot will compare males in control group and
males exposed to 4C for 48 hours. This group is called 'coma'. The second plot
will compare males who were exposed to 45C and either hit heat stupor (hs) or
were just exposed to heat and did not hit heat stupor (heat). 
```{r boxplots of viability}
#cold treatment boxplot
cvia1$treat<- factor(cvia1$treat, levels= c("Control", "Cold"))
p1 <- cvia1 %>%   
  ggplot(aes(x=treat, y=via, fill=treat)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  scale_fill_manual(values=c("white","light blue")) +
  labs(x="Treatment", y="% Viability") +
  labs(fill='Treatments') +
  ylim(0,100)+
  geom_jitter(position = position_jitterdodge()) 

p1 

#cold treatment boxplot with y axis log scale to see if data is better
#visualized
p2 <- cvia1 %>%   
  ggplot(aes(x=treat, y=via, fill=treat)) +
  geom_boxplot(outlier.shape = NA) +
  theme_classic() +
  scale_fill_manual(values=c("white","light blue")) +
  labs(x="Treatment", y="% Viability") +
  labs(fill='Treatments') +
  scale_y_continuous(trans='log10') +
  geom_jitter(position = position_jitterdodge()) 
p2 

#heat treatment boxplot
p3 <- hvia1 %>%   
 ggplot(aes(x=treat, y=via, fill=treat)) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values=c("white","red", "dark red")) +
  theme_classic() +
  labs(x="Treatment", y="% Viability") +
  labs(fill='Treatments') +  
  ylim(0,100)+
  geom_jitter(position = position_jitterdodge())

p3

#heat treatment boxplot with y axis log scale to see if data is better 
#visualized
p4 <- hvia1 %>%   
 ggplot(aes(x=treat, y=via, fill=treat)) +
  geom_boxplot(outlier.shape = NA) +
  scale_fill_manual(values=c("white","red", "dark red")) +
  theme_classic() +
  labs(x="Treatment", y="% Viability") +
  labs(fill='Treatments') +  
  scale_y_continuous(trans='log10') +
  geom_jitter(position = position_jitterdodge())

p4

p1+p3

```
These graphs are showing that sperm count is effected by temperature compared 
to the control for both heat and cold treatments. More statistical testing is needed to 
determine if these treatment cause a significant decrease. 



# 6. Plotting Viability in both datasets with violin plots
Another way to visualize this dataset could be using violin plots. 
```{r violin plots of viability}

#cold treatment violin plot with points
cvia1$treat<- factor(cvia1$treat, levels= c("Control", "Cold"))
h1<- cvia1 %>%   
 ggplot(aes(x=treat, y=via, fill=treat)) +
    geom_violin(width=1) +
  geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5)+
    theme_classic() +
    scale_fill_manual(values=c("white", "light blue")) +
    labs(x="Treatment", y="% Viability") +
    ylim(0,100)+
    labs(fill='Treatments')
h1

#heat treatment violin plot with points
h2 <- hvia1 %>%   
 ggplot(aes(x=treat, y=via, fill=treat)) +
    geom_violin(width=1) +
    geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5)+
    theme_classic() +
    scale_fill_manual(values=c("white","red", "dark red")) +
    labs(x="Treatment", y="% Viability") +
    ylim(0,100)+
    labs(fill='Treatments')
h2

h1+h2

```
These graphs are just showing another way to visualize how temperature decreases
sperm viability in males. 

# 8. Feedback Implementation 
After receiving feedback from my peers I implemented changes to my final
project. I made some minor changes such as making sure my words did not run
off the page, italicizing scientific names, mute loading packages, and changed 
the order of cold treatment graphs so control is first. I also added summaries 
below each plot of findings, added points to my violin plots, used patchwork to 
combine the best two graphs together, renamed treatments, and adjusted y axis.



# 7. References

<div id="refs"></div>


